Spatiotemporal Multi-Resolution Approximation of the Amari Type Neural Field Model

### 2012-01-16 Multiresolution_framework/ltx/manuscript.tex

Neural field models describe the mass action of the central nervous system and are a critical link in our understanding of the biophysics of the EEG. These models typically describe the macroscopic dynamics of the human brain, but are also descriptive of finer-scale neurodynamics. For example.. integrate and fire... The utility in these models to answer questions in neuroscience lies in the ability to infer parameters from data. The ability to create patient-specific neural models will contribute to our knowledge base of diseases such as epilepsy and will enable the development of new treatment strategies. This is particularly relevant to the advent of new devices that use therapeutic electrical stimulation. Stimulation strategies for devices currently operate in an open loop, where stimulation parameters are chosen using a process of trial and error. Therefore, there exists enormous potential to improve the performance of these devices using systems theory, which in turn requires a suitable dynamic model. Neural mass and neural field models are ideal candidates for estimation algorithms due to their strong links with physiology and their parsimony. It is expected that parameters of the neural fields models will be patient-specific and will, therefore, need to be inferred from data.

The first work describing data-driven mesoscopic neural modeling used a neural mass model to fit EEG data~\cite{Valdes1999}. This approach was extended to coupled neural masses through a Bayesian estimation scheme dubbed Dynamic Causal Modeling (DCM)~\cite{David2003}. Following this work, data-driven modeling was extended to continuum field equations that explained the richer dynamics of spatiotemporal neural fields \cite{Galka2008,schiff2008kalman,Daunizeau2009}. Most recently, a framework was developed where a finite element model (FEM) of the neural field (via a global Galerkin projection) was formed, using a basis function decomposition, to transform the PDE neural field equations into a finite dimension system to facilitate efficient state and parameter estimation~\cite{Freestone2011}. This paper is an extension to the aforementioned study, where we derive a neural field state-space model that accounts for the multi-resolution architecture and spatial dynamics of the human cortex. In this way, a flexible framework is created, whereby both macroscopic and microscopic behavior of the system can be represented simultaneously \dean{and completely}.

@@

\section{IDE Neural Field Model}
The stochastic IDE form of the Amari neural field  formulation~\cite{Amari1977} is given by (see~\cite{Freestone2011} for a full derivation)
\begin{equation}\label{eq:DiscreteTimeModel}
	v_{t+1}\left(\mathbf{r}\right) = 
	\xi v_t\left(\mathbf{r}\right) + 
	T_s \int_\Omega { 
	    w\left(\mathbf{r},\mathbf{r'}\right)
	    f\left(v_t\left(\mathbf{r}'\right)\right) 
	\, d\mathbf{r}'} 
	+ e_t\left(\mathbf{r}\right), 
\end{equation}
where the post-synaptic membrane voltage at time $t$ of a population of neurons at position $\mathbf r$ is denoted $v_t\left(\mathbf r\right)$. Synaptic dynamics are included through the parameter $\xi=1-T_s/\tau$, where $\tau$ is the synaptic time constant and $T_s$ is the sampling time. The connectivity strength between neural populations at a distance $\mid\mathbf{r}-\mathbf{r'}\mid$ is described by the connectivity kernel $w\left(\mathbf{r}-\mathbf{r}'\right)$. The connectivity kernel is taken as a ``Mexican hat'' function, which describes local excitation and lateral inhibition \cite{Amari1977,Atay2005}. The term $e_t(\mathbf r)$ is a zero mean Gaussian disturbance, spatially coloured but temporally white, with covarianve function 
\begin{equation}
cov\left(e_{t}\left(\mathbf{r}\right),e_{t+t'}\left(\mathbf{r'}\right)\right)=\eta(\mathbf{r}-\mathbf{r'})\delta(t-t'),
\label{eq:FieldDisturbance}
\end{equation}
where $\delta(\cdot)$ is the Dirac-delta function. The firing rate of the presynaptic neurons is related to the post-synaptic membrane potential by the activation function $f(v_t(\mathbf{r})) = \varsigma v_t(\mathbf{r})$ \cite{VanRotterdam1982,Murphy2009}.

The observation equation describing the intracranial electrophysiological recordings is given by
\begin{equation}\label{eq:ObservationEquation}
	y_t(\mathbf{r}_{n_y}) = \int_{\Omega} { m\left(\mathbf{r}_{n_y}-\mathbf{r}'\right) v_t\left(\mathbf{r}'\right) \, d\mathbf{r}'} + \boldsymbol\epsilon_t(\mathbf{r}_{n_y}), 
\end{equation}
where $m\left(\mathbf{r}_{n_y}-\mathbf{r}'\right)$ is the observation kernel at location $\mathbf{r}_{n_y}$ and  $\boldsymbol{\epsilon}_{t}\sim \mathcal{N}\left(\mathbf{0},\mathbf{\Sigma}_{\epsilon}\right)$  is an i.i.d. Gaussian white noise process and the covariance matrix $\mathbf{\Sigma}_{\epsilon}=\sigma^2_{\epsilon}\mathbf I_{n_y}$, where $\mathbf I$ denotes the identity matrix. 
\section{MRA of the IDE in State-Space}
The multi-resolution approximation \cite{Mallat1989a} of the neural IDE (MRAIDE) is obtained by decomposing both the field, $v_t(\cdot)$, and the connectivity kernel, $w(\cdot)$, (assuming square-integrable functions) using translations and dilations of a scaling function $\phi(r)$ and a mother wavelet $\psi(r)$. Considering a one-dimensional field, the connectivity kernel is decomposed as,
\begin{equation}
 w\left(r-r'\right)=\sum_{l \in \mathbb{Z}}\alpha_{j_0,l}\phi_{j_0,l}\left(r-r'\right)+\sum_{j\geq j_0}^{\infty} \sum_{l \in \mathbb{Z}}\beta_{j,l}\psi_{j,l}\left(r-r'\right), 
\label{eq:KernelExpansion}
\end{equation}
where $\alpha_{j_0,l}$ are the approximation coefficients at the lowest scale $j_0$, and $\beta_{j,l}$ are the detail coefficients at different scales $j$, with $\phi_{j,l}\left(r\right)=2^{\frac{j}{2}}\phi\left(2^jr-l\right) $ and $\psi_{j,l}\left(r\right)=2^{\frac{j}{2}}\psi\left(2^jr-l\right)$. Integers $j$ and $l$ are the scale and translation parameters, respectively. Scaling functions retain the lowest frequency components, while the wavelet functions extract successively higher frequency components due to their low-pass and band-pass properties, respectively. The field is decomposed using
\begin{equation}
 v_t\left(r\right)=\sum_{l \in \mathbb{Z}}x_{t,j_{0},l}\phi_{j_{0},l}\left(r\right)+\sum_{j\geq j_0}^{\infty} \sum_{l \in \mathbb{Z}} \check{x}_{t,j,l}\psi_{j,l}\left(r\right),
\label{eq:FieldExpansion}
\end{equation}
where $x_{t,j_{0},l}$ and $\check{x}_{t,j,l} $ are the coefficients of the expansion and constitute the state vector at time $t$.

@@

>>>>

Neural field models describe the mass action of the central nervous system and are a critical link in our understanding of the biophysics of the EEG. The key components of these models typically describe the macroscopic dynamics of the human brain, but can also descriptive of finer-scale neurodynamics. For example, the post-synaptic response kernels (the pulse to wave function) can be descriptive of inhibitory and excitatory post-synaptic potentials in microscopic single neural models, such as integrate and fire, or an ensemble response to stimulation in macroscopic neural mass models. Furthermore, the activation function (wave to pulse function) can be used used to model spiking statistics across spatial scales from single neurons (where is can be considered a CDF of the probability of firing) to mean firing rates of neural masses. The key factor linking the key functions of the model to particular spatial scales is geometry of the connectivity, hence the ability to infer a multi-resolution connectivity kernel will enable to the formulation of a multi-resolution neural field model.

The utility of these models for answering questions in clinical neuroscience and neurology lies in the ability the infer parameters from electrophysiological data. The ability to create patient-specific neural models will contribute to our knowledge base of diseases such as epilepsy and will enable the development of new treatment strategies. This is particularly relevant to the advent of new devices that use therapeutic electrical stimulation. Stimulation strategies for devices currently operate in an open loop, where stimulation parameters are chosen using a process of trial and error. Therefore, there exists enormous potential to improve the performance of these devices using systems theory, which in turn requires a suitable dynamic model. Neural mass and neural field models are ideal candidates for estimation algorithms due to their strong links with physiology and their parsimony. It is expected that parameters of the neural fields models will be patient-specific and will, therefore, need to be inferred from data.

The first work describing data-driven mesoscopic neural modelling used a neural mass model to fit EEG data~\cite{Valdes1999}. This approach was extended to coupled neural masses through a Bayesian estimation scheme dubbed Dynamic Causal Modelling (DCM)~\cite{David2003}. Following this work, data-driven modelling was extended to continuum field equations that explained the richer dynamics of spatiotemporal neural fields \cite{Galka2008,schiff2008kalman,Daunizeau2009}. Most recently, a framework was developed where a finite element model (FEM) of the neural field (via a global Galerkin projection) was formed, using a basis function decomposition, to transform the PDE neural field equations into a finite dimension system to facilitate efficient state and parameter estimation~\cite{Freestone2011}. This paper is an extension to the aforementioned study, where we derive a neural field state-space model that accounts for the multi-resolution architecture and spatial dynamics of the human cortex. In this way, a flexible framework is created, whereby both macroscopic and microscopic behaviour of the system can be represented simultaneously \dean{and completely}.

@@

\section{IDE Neural Field Model}
The stochastic IDE form of the Amari neural field  formulation~\cite{Amari1977} is given by (see~\cite{Freestone2011} for a full derivation)
\begin{equation}\label{eq:DiscreteTimeModel}
	v_{t+1}\left(\mathbf{r}\right) = 
	\xi v_t\left(\mathbf{r}\right) + 
	T_s \int_\Omega { 
	    w\left(\mathbf{r},\mathbf{r'}\right)
	    f\left(v_t\left(\mathbf{r}'\right)\right) 
	\, d\mathbf{r}'} 
	+ e_t\left(\mathbf{r}\right), 
\end{equation}
where the post-synaptic membrane voltage at time $t$ of a population of neurons at position $\mathbf r$ is denoted $v_t\left(\mathbf r\right)$. Membrane dynamics are included through the parameter $\xi=1-T_s/\tau$, where $\tau$ is the membrane time constant and $T_s$ is the sampling time. The connectivity strength between neural populations at a distance $\mid\mathbf{r}-\mathbf{r'}\mid$ is described by the connectivity kernel $w\left(\mathbf{r}-\mathbf{r}'\right)$. The connectivity kernel is taken as a ``Mexican hat'' function, which describes local excitation and lateral inhibition \cite{Amari1977,Atay2005}. The term $e_t(\mathbf r)$ is a zero mean Gaussian disturbance, spatially coloured but temporally white, with covariance function 
\begin{equation}
cov\left(e_{t}\left(\mathbf{r}\right),e_{t+t'}\left(\mathbf{r'}\right)\right)=\eta(\mathbf{r}-\mathbf{r'})\delta(t-t'),
\label{eq:FieldDisturbance}
\end{equation}
where $\delta(\cdot)$ is the Dirac-delta function. The firing rate of the presynaptic neurons is related to the post-synaptic membrane potential by the activation function $f(v_t(\mathbf{r})) = \varsigma v_t(\mathbf{r})$ \cite{VanRotterdam1982,Murphy2009}.

The observation equation describing the intracranial electrophysiological recordings is given by
\begin{equation}\label{eq:ObservationEquation}
	y_t(\mathbf{r}_{n_y}) = \int_{\Omega} { m\left(\mathbf{r}_{n_y}-\mathbf{r}'\right) v_t\left(\mathbf{r}'\right) \, d\mathbf{r}'} + \boldsymbol\epsilon_t(\mathbf{r}_{n_y}), 
\end{equation}
where $m\left(\mathbf{r}_{n_y}-\mathbf{r}'\right)$ is the observation kernel at location $\mathbf{r}_{n_y}$ and  $\boldsymbol{\epsilon}_{t}\sim \mathcal{N}\left(\mathbf{0},\mathbf{\Sigma}_{\epsilon}\right)$  is an i.i.d. Gaussian white noise process and the covariance matrix $\mathbf{\Sigma}_{\epsilon}=\sigma^2_{\epsilon}\mathbf I_{n_y}$, where $\mathbf I$ denotes the identity matrix.

\section{MRA of the IDE in State-Space}
The multi-resolution approximation \cite{Mallat1989a} of the neural IDE (MRAIDE) is obtained by decomposing both the field, $v_t(\cdot)$, and the connectivity kernel, $w(\cdot)$, (assuming square-integrable functions) using translations and dilations of a scaling function $\phi(r)$ and a mother wavelet $\psi(r)$. Considering a one-dimensional field, the connectivity kernel is decomposed as,
\begin{equation}
 w\left(r-r'\right)=\sum_{l \in \mathbb{Z}}\alpha_{j_0,l}\phi_{j_0,l}\left(r-r'\right)+\sum_{j\geq j_0}^{\infty} \sum_{l \in \mathbb{Z}}\beta_{j,l}\psi_{j,l}\left(r-r'\right), 
\label{eq:KernelExpansion}
\end{equation}
where $\alpha_{j_0,l}$ are the approximation coefficients at the lowest scale $j_0$, and $\beta_{j,l}$ are the detail coefficients at different scales $j$, with $\phi_{j,l}\left(r\right)=2^{\frac{j}{2}}\phi\left(2^jr-l\right) $ and $\psi_{j,l}\left(r\right)=2^{\frac{j}{2}}\psi\left(2^jr-l\right)$. Integers $j$ and $l$ are the scale and translation parameters, respectively. Scaling functions retain the lowest frequency components, while the wavelet functions extract successively higher frequency components due to their low-pass and band-pass properties, respectively. The field is decomposed using
\begin{equation}
 v_t\left(r\right)=\sum_{l \in \mathbb{Z}}x_{t,j_{0},l}\phi_{j_{0},l}\left(r\right)+\sum_{j\geq j_0}^{\infty} \sum_{l \in \mathbb{Z}} \check{x}_{t,j,l}\psi_{j,l}\left(r\right),
\label{eq:FieldExpansion}
\end{equation}
where $x_{t,j_{0},l}$ and $\check{x}_{t,j,l} $ are the coefficients of the expansion and constitute the state vector at time $t$. \dean{The decomposition enables a separation of the spatial characteristics, which are considered static functions, that are weighted by dynamic coefficients.  }

@@

The goal of this section is to demonstrate how to formulate a state-space model in a canonical form such that standard algorithms may be applied to estimate the multi-resolution connectivity structure. Therefore, the goal is to form a parameterised equation for coefficients of the field basis functions $\mathbf{x}_t$, which becomes the state vector.
